### 2014-12-10 doc/thesis/typedlua.tex
Adding a section about interfaces and type aliases to chapter 3

\section{Tables and Interfaces}
\label{sec:tables}

@@

Typed Lua includes \emph{interfaces} as syntactic sugar to table types
that specify records, as writing table types can be unwieldy when
records get bigger and types of record fields get more complicated.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

@@

@@

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of an interface,
userdata, or type alias.
As we saw, an interface declaration is just a shortcut to a table type
that represents a record.
An userdata declaration is similar to an interface declaration,
but it also includes its name as the \emph{brand} of the table type.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two different userdata are not subtype
of each other because they do not share the same brand.
A type alias declaration assigns a name to a first-level type.
Even though the type alias declaration may make the interface declaration
look redundant, it does not have a syntax as nice as the syntax
of the interface declaration for defining records.
The type alias declaration is specially useful when we need to create
names for types that are not records.

>>>>

\section{Tables}
\label{sec:tables}

@@

\section{Interfaces and type aliases}
\label{sec:alias}

Typed Lua includes \emph{interfaces} as syntactic sugar to table types
that specify records, as writing table types can be unwieldy when
records get bigger and types of record fields get more complicated.
For instance, we can use the following interface as an alias to the
type of the record we showed in the last section:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

@@

Typed Lua also includes a way to create aliases for types that does not
follow the records syntax.
However, we limitim aliases to union types, function types, table types,
and recursive types.
The reason for this restriction is to avoid the creation of aliases
to predefined types.

For instance, we can use the \texttt{typealias} construct to define an
alias to the type of the abstract syntax tree for a language of simple
arithmetic expressions:
\begin{verbatim}
    local typealias Exp = {"tag":"Number", 1:number}
                        | {"tag":"Add", 1:Exp, 2:Exp}
                        | {"tag":"Sub", 1:Exp, 2:Exp}
                        | {"tag":"Mul", 1:Exp, 2:Exp}
                        | {"tag":"Div", 1:Exp, 2:Exp}
\end{verbatim}

The type \texttt{Exp} is a recursive type that resembles the
algebraic data types from functional programming.
Typed Lua also includes typing rules that let programmers
deconstruct unions of table types, and it resembles
pattern matching from functional programming.
We can use this feature to write an evaluation function for
our simple language:
\begin{verbatim}
    local function eval (e:Exp):number
      if e.tag == "Number" then return e[1]
      elseif e.tag == "Add" then return eval(e[1]) + eval(e[2])
      elseif e.tag == "Sub" then return eval(e[1]) - eval(e[2])
      elseif e.tag == "Mul" then return eval(e[1]) * eval(e[2])
      elseif e.tag == "Div" then return eval(e[1]) / eval(e[2])
      end
    end
\end{verbatim}

When Typed Lua finds the pattern \texttt{n.s1 == s2}, it checks
if the type of the variable \texttt{n} is a union of table types
and if \texttt{s1} and \texttt{s2} are strings.
If that is the case, it constrains the type of \texttt{n} inside
the \texttt{if} to the table type that has the pair \texttt{s1:s2}.
For instance, in the case of \texttt{e.tag == "Add"}, Typed Lua
constrains the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
This is safe because Typed Lua restricts this pattern to literal types only.

Even though the type alias declaration may make the interface declaration
look redundant, it does not have a syntax as nice as the syntax
of the interface declaration for defining records.
For this reason worth keeping two different constructs.
One for records specifically and another to more general types.

@@

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of an interface,
userdata, or type alias.
As we saw, an interface declaration is just a shortcut to a table type
that represents a record.
An userdata declaration is similar to an interface declaration,
but it also includes its name as the \emph{brand} of the table type.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two different userdata are not subtype
of each other because they do not share the same brand.
A type alias declaration assigns a name to a first-level type.
